---
title: "Introduction to R"
author: "P. Adames"
date: "11/14/2020"
bibliography: bib.bib
output: 
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 4
    highlight: tango
  urlcolor: blue
  linkcolor: red
  citecolor: purple
  citation_package: natbib
  latex_engine: lualatex
  includes:
    in_header: preamble.tex
documentclass: article
classoption: oneside
fontsize: 11
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r loading packages, echo=FALSE}
# from https://www.r-bloggers.com/2015/11/creating-nice-tables-using-r-markdown/

# List of packages required for this analysis
pkg <- c("dplyr", "knitr", "devtools", "DT", "xtable")
# Check if packages are not installed and assign the
# names of the packages not installed to the variable new.pkg
new.pkg <- pkg[!(pkg %in% installed.packages())]
# If there are any packages in the list that aren't installed,
# install them
if (length(new.pkg)) {
  install.packages(new.pkg, repos = "http://cran.rstudio.com")
}
# Load the packages into R
library(dplyr)
library(knitr)
library(DT)
library(xtable)
```

# Introduction to R

The introduction to R document in the official repository of packages for the R language has 105 pages and 14 sections (<https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf>). It covers data structures, reading and writing (I/O) to files, probability distributions, statistical models, conditional statements and other forms of controlling the execution of a program,  user functions, graphics, extension via packages, and commands to interact with the host operating system.

Undoubably too much for a 45 minute session. Therefore the focus will be on two aspects:

 - language specific syntax
 - basic I/O and interoperability with the operating system

These two aspects can give the student a quick sense of usability for the R langauage. This approrach leaves an essential strengtgh of R for later, its nature as a statistical computing and graphics language.


# Language Syntax

Programing languages model information using data structures. At its core, R represents everything  either as an  object or a function. There are two immediate consequences from this: there no absolute need to use for loops ---although they exist--- because functions are designed to work very effectively on vectorized data structures, and scalars only exist as unidimensional data structures of length 1 [@IhakaGentleman96].

We will emphasize these two points to set the stage for elegant and fast R code. R was a functional language long before this became a buzz word in Computer Science.

## Data structures

R uses the data structures illustrated in the table below from [@WickhamHadley2015AR, p.13].
Vectors, matrices, adn arrays can store values of the same data type, whether they are are numeric, integer, logical, or character. 
In contrast, lists and data frames can store mixed data types.
For 3d and higher dimensionality the native data structure is an object of class _array_.

```{r data structures, echo=FALSE, results='asis'}
Homogeneous <- c("Atomic vector", "Matrix", "Array")
Heterogeneous <- c("List", "Data frame", "")

kable( data.frame(Homogeneous, Heterogeneous, 
                 row.names = c("1d", "2d", "nd"), 
                 stringsAsFactors = FALSE))#,
       # caption = 'Native data structures in R according to the data type they can store and the number of dimensions they use.')
```

Because every data in R is an object, all of these structures have attributes like name, dimension, and a \texttt{class}. 
They can also receive and carry user-defined attributes.


## Data types

 - *numeric:* double precision floating point real numbers.
 - *integer:* model integer numbers.
 - *logical:* can be TRUE, T, 1 or FALSE, F, 0.
 - *character:* models aa text character in an encoding.

## Built in functions

The magic of the generic dispatch mechanism of S3 R objects.


### Iteration over data structures


### I/O


### Interoperatibility with the host operating system


## User-functions


## Execution control


# References







